### 2015-04-23 papers/histogram/thesis/thesis.tex
some updates to histogram thesis

\renewcommand{\title}{Optimizing Monte Carlo state space exploration
  of the square-well fluid}
\renewcommand{\author}{Michael A. Perlin}
\newcommand{\institution}{Oregon State University}
\newcommand{\department}{Department of Physics}
\newcommand{\supervisor}{Dr. David Roundy}
\renewcommand{\date}{19 April 2015}

@@

We wish to identify and develop efficient numerical methods for
determining thermodynamic properties of the square-well fluid, so as
to test square-well density functional theories. The liquid-vapor
phase interface is an interesting regime for testing density
functional theories, but unbiased Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low (i.e. liquid-state) temperatures. Several
generic Monte Carlo ``histogram methods'' for collecting statistics on
low energy system states have been developed, but little or no
literature exists on their systematic comparison and their application
to the square-well fluid. The square-well fluid in particular
introduces application challenges not manifest in traditional models
for testing and benchmarking such numerical techniques (e.g. the Ising
model). We propose to implement these methods in Monte Carlo
simulations of the square-well fluid, determine appropriate
performance metrics for their comparison, and identify those methods
which are most efficient and effective at sampling otherwise
improbable system states.

@@

A major challenge in the field of classical density functional theory
is understanding critical point behavior of fluids, the study of which
via simulations requires exploring the state space boundary between
liquid and gaseous fluid phases. For a generic square-well fluid to be
observed in a liquid state, the fluid must have a relatively localized
distribution of the spheres which make up the fluid. Such a
distribution implies a low energy fluid state; conversely, a low
energy liquid state of a square-well fluid otherwise observable in
both liquid and gaseous phases implies a localized distribution of
spheres. The search for liquid states is thus equivalent to the search
for low energy states. In unbiased Monte Carlo simulations, however,
highly localized distributions of spheres are extremely unlikely to
occur, making such simulations impractical for studying the
square-well fluid.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/square-well.pdf}
  \caption[The square-well pair potential]{The square-well pair
    potential can be used to model short-range attractive forces to
    first order. The infinite potential at $r<\sigma$ simply enforces
    the condition that spheres cannot overlap. At $r>\lambda\sigma$,
    the potential is null, and makes no contribution to the fluid's
    internal energy.}
  \label{fig:pair_potential}
\end{figure}

The square-well (SW) fluid is a model used in classical density
functional theories to capture low order effects of short-range
attractive forces, such as the van der Waals force. The fluid is
composed of spheres with diameter $\sigma$ which have a pair potential
\begin{align}
  v_{sw}\p{\vec r}=\left\{
    \begin{array}{ll}
      \infty & \abs{\vec r}<\sigma \\
      -\epsilon & \sigma<\abs{\vec r}<\lambda\sigma \\
      0 & \abs{\vec r}>\lambda\sigma
    \end{array}
  \right., \label{eq:pair_potential}
\end{align}
also shown graphically in Figure \ref{fig:pair_potential}, where the
parameters $\lambda$ and $\epsilon$ are referred to as the well width
and depth, respectively. The first ($\abs{\vec r}<\sigma$) part of
this potential forbids spheres from overlapping, whereas the second
($\sigma<\abs{\vec r}<\lambda\sigma$) associates an energy $-\epsilon$
with each pair of spheres whose centers are within distance of
$\lambda\sigma$ of each other (where typically
$\lambda\in\sp{1.05,3}$). The net potential energy of the square-well
fluid is thus
\begin{align}
  E=\sum_{i<j}v_{sw}\p{\vec r_i-\vec r_j},
  \label{eq:internal_energy}
\end{align}
where $\vec r_i$ is the position of the $i$-th sphere. As the
potential energy $E$ is the primary form of energy concerning us in
this paper, we will refer to $E$ as simply the ``energy'' of the
fluid. An important feature of the square-well fluid is that its
energy is always an integer multiple of the well depth. A homogeneous
square-well fluid is thus uniquely identified by its well width
$\lambda$ and filling fraction $\eta$ (i.e. the proportion of space
filled by spheres; a dimensionless density), as all other properties
can be normalized to the natural energy scale $\epsilon$ and length
scale $\sigma$. In practice, our simulation codes use dimensionless
energies $E/\epsilon$, temperatures $kT/\epsilon$, and distances
$r/\sigma$.

@@

While model systems are powerful tools for understanding complex
physical systems, they do not themselves exist in the real
world. Consequently, direct experimental tests of model system
theories (e.g. square-well density functional theories) are not
possible. For this reason, model system theories are commonly tested
against Monte Carlo simulations. Proper implementation of Monte Carlo
methods to study completely characterized systems ensures that
statistical results from simulations converge on the exact properties
of the simulated system in the infinite simulation time
limit\fixme{cite?}. Furthermore, uncertainties in quantities computed
via Monte Carlo simulations are typically well-defined, monotonically
decreasing functions of simulation time, allowing one to run
simulations to the desired level of accuracy.

@@

Algorithm \ref{alg:MC} provides a sketch of unbiased Monte Carlo fluid
simulations. Such an algorithm is ``unbiased'' in the sense that it
collects statistics (i.e. data) on all valid system
configurations. Statistics whose collection time scales as $\mathcal
O\p{1}$ with system size can be collected after every move (defined in
the algorithm), while statistics whose collection time scales as
$\mathcal O\p{N}$, where $N$ is the number of simulated elements
(e.g. spheres), can be collected after each iteration. In general,
collection with $\mathcal O\p{\chi}$ time scaling in $N$ should not
occur more often than once every $\chi$ moves \fixme{explain better?},
so as to ensure that scaling up simulations does not cause them to
asymptotically spend all of their time collecting statistics, and no
time actually simulating the fluid. Collected statistics determine
thermodynamic properties of the simulated fluid.

In this work, we are concerned with simulating the homogeneous
square-well fluid. To avoid edge effects resulting from fluid behavior
near a wall, we enforce periodic boundary conditions. These boundary
conditions suppress all density fluctuations on scales larger than the
dimensions of the simulated fluid cell, thereby introducing a source
of error; addressing this error, however, is outside the scope of this
work. \fixme{Anything else to add?}

@@

Due to the fact that unbiased Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ is directly proportional to
the density of states $D\p{X}$ in $X$ \fixme{explain?}. It is
sometimes the case, however, that the density of states in some region
$R$ in the range of possible $X$ is so low that it is practically
impossible to sufficiently sample $R$, that is, accumulate a
statistically significant $H\p{R}$, in any reasonable amount of time.

@@

Histogram methods provide a means to address unbiased Monte Carlo's
inability to sufficiently sample regions of parameter space with a low
density of states by introducing a bias into the otherwise random
sampling of state space. A weighting function $w$ is introduced, whose
domain is the value of some property $X\p{s}$ which depends on the
system state $s$. An additional condition is then added to step
\ref{alg:MC_move} of Algorithm \ref{alg:MC} in order to accept an
attempted move: the weights $w\sp{X\p{s_i}}$ and $w\sp{X\p{s_f}}$ of
the initial (pre-move) state $s_i$ and final (post-move) state $s_f$
are used to determine the probability $P\p{s_i\to s_f}$ of accepting
an otherwise valid move via
\begin{align}
  P\p{s_i\to s_f}=\min\set{\f{w\sp{X\p{s_f}}}{w\sp{X\p{s_i}}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\sp{X\p{s_f}}>w\sp{X\p{s_i}}$, the move
$s_i\to s_f$ is accepted; when $w\sp{X\p{s_f}}<w\sp{X\p{s_i}}$, the
ratio of these weights determines the probability of accepting the
move. Due to the fact that only ratios of weights determine
$P\p{s_i\to s_f}$, the weights $w\sp{X\p{s}}$ are scale-invariant,
meaning that their effect on simulations is unchanged by scale factors
that are constant with respect $X\p{s}$.

@@

Employing biased Monte Carlo simulations, sketched out in Algorithm
\ref{alg:biased_MC}, allows one to construct weight functions that
favor some region of state space over others, as transitions to states
with higher weights are always accepted, whereas transitions to states
with lower weights may be rejected, artificially preventing the
simulated system from leaving interesting regions of state
space. Crucially, the bias introduced by weights can be reversed when
computing system properties from sampling statistics, as the weight of
a particular state is directly proportional to the probability bias of
that state; that is, a state with a weight of 2 will be sampled twice
as often as it would have been with a weight of 1 (\fixme{include a
  subsection on detailed balance? cite?}). If we wish to convert a
histogram $H\p{X}$ of observations in a biased Monte Carlo simulation
into a numerical unnormalized density of states $\tilde D\p{X}$ in
$X$, we therefore point-wise divide the histogram by the corresponding
weights $w\p{X}$, i.e.
\begin{align}
  \tilde D\p{X}=\f{H\p{X}}{w\p{X}}.
  \label{eq:dos_unnormed}
\end{align}
The normalized density of states $D\p{X}$ in $X$ is then
\begin{align}
  D\p{X}=\f{\tilde D\p{X}}{\sum_X\tilde D\p{X}}
  =\f{H\p{X}/w\p{X}}{\sum_XH\p{X}/w\p{X}}.
  \label{eq:dos}
\end{align}
In general, the domain and shape of the weight function will depend on
the desired yields (e.g.  density of states, heat capacity) of a
simulation. A ``histogram method'' is an algorithm for determining a
weight function appropriate for a particular simulation.

@@

The sort of Monte Carlo simulations described in this paper are only
capable of collecting statistics on system microstates, and functions
of microstates. For example, in simulation one might periodically
compute both the energy $E\p{s}$ and some system property $X\p{s}$,
both of which are determined by the system microstate $s$, in order to
find the mean value of $X$ at any given energy $E$, i.e. $X\p{E}$. In
the real world, however, information about a system's microstate and
functions of microstates (e.g. the energy $E$) is generally is
inaccessible. One therefore cannot measure $X\p{E}$ directly. Instead,
one typically measures the dependence of thermodynamic properties on
macroscopic state variables such as temperature, i.e. $X\p{T}$.

The temperature dependence $X\p{T}$ can be expressed in terms of the
density of states $D\p{E}$ and energy dependence $X\p{E}$ by
\begin{align}
  X\p{T}=\f1{Z\p{T}}\sum_EX\p{E}D\p{E}e^{-E/kT},
  \label{eq:temp_dep_X_norm}
\end{align}
where the partition function $Z\p{T}$ is
\begin{align}
  Z\p{T}=\sum_ED\p{E}e^{-E/kT}.
\end{align}
To reduce redundant computations and numerical error in
implementation, we will use an unnormalized partition function
\begin{align}
  \tilde Z\p{T}=\sum_E\tilde D\p{E}e^{-E/kT},
\end{align}
in terms of which
\begin{align}
  X\p{T}=\f1{\tilde Z\p{T}}\sum_EX\p{E}\tilde D\p{E}e^{-E/kT}.
  \label{eq:temp_dep_X}
\end{align}
Sufficient accumulation of statistics on $X\p{E}$ at all available
energies thus in principle allows one to determine $X\p{T}$ for any
$T$. In practice, the density of states can fall off so quickly with
energy that some range of allowable energies is practically
inaccessible via Monte Carlo simulations, biased or otherwise. In such
a case, computing $X\p{T}$ to a reasonable degree of accuracy requires
sufficiently sampling the energies at which $\tilde D\p{E}e^{-E/kT}$
dominates the sum in (\ref{eq:temp_dep_X_norm}).

@@

% \subsubsection{Detailed balance}
% \label{sec:detailed_balance}

% \fixme{In this section I will explain why and how weights work as
% advertised. Essentially, we will consider the effect of weights on a
% two-state system (or on the transitions between two particular
% states in an otherwise multi-state system), and convince ourselves
% that the biasing effect of weights is the identical in an ensemble
% of transitions between any number of states.}

@@

A standard construction of the weight array $w\p{E}$ for what is
called a canonical Monte Carlo simulation involves \fixme{better word}
choosing a particular temperature $T_0$, and using weights
proportional to the Boltzmann distribution at that temperature, i.e.
\begin{align}
  w\p{E}=e^{-E/kT_0},
\end{align}
where there is no reason to normalize $w\p{E}$ due to the fact that
only ratios of weights, as per (\ref{eq:move_prob}), are ever used in
simulation. The fact that only ratios of weights are used in
simulation makes $w\p{E}$ scale-invariant.

The partition function at $T=T_0$ for simulations with canonical
weights is
\begin{align}
  \tilde Z\p{T_0}=\sum_E\tilde D\p{E}e^{-E/kT}
  =\sum_EH\p{E}/w\p{E}~e^{-E/kT} =\sum_EH\p{E},
  \label{Z_canonical}
\end{align}
and the value of a thermodynamic property $X\p{T_0}$
\begin{align}
  X\p{T_0}=\f1{\tilde Z\p{T_0}}\sum_EX\p{E}\tilde D\p{E}e^{-E/kT_0}
  =\f{\sum_EX\p{E}H\p{E}}{\sum_EH\p{E}}.
  \label{eq:X_canonical}
\end{align}
The simplifications in in (\ref{Z_canonical}) and
(\ref{eq:X_canonical}), which have no explicit dependence on $T_0$,
occur because canonical Monte Carlo simulations sample energies in
proportion to the distribution (over energy) of microstates at a
temperature of $T_0$.  Canonical Monte Carlo simulations can therefore
fail to sufficiently sample energies which are important
(i.e. energies with a non-negligible state probability density) at
different temperatures. As a consequence, such simulations should not
be used to determine properties $X\p{T\ne T_0}$, and are thus referred
to as ``fixed temperature'' simulations.

Though canonical Monte Carlo is simple to implement, its inability to
investigate a system at more than one temperature at a time is a
disadvantage for determining the temperature dependence of system
properties. In order to find the behavior of $X\p{T}$, one must run
many simulations at discrete temperature intervals; each such
simulation will yield one sample (e.g. one data point) of $X\p{T}$.

@@

For all but the most trivial or well-studied systems, however, the
density of states $D\p{E}$ is not known prior to simulation, and is in
fact one of the system properties which a simulation is intended to
determine. Nonetheless, this toy method has pedagogical value by
enabling discussion of ``optimal'' weights for generating a flat
histogram. All algorithms to construct $w\p{E}$ with the aim of
producing a flat histogram should therefore converge on the same
weight array, given by (\ref{eq:flat_weights}).

@@

\begin{algorithm}[H]
  \caption{Finding weights by optimizing energy walker currents}
  \label{alg:optimized_ensemble}
\end{algorithm}

\subsection{The transition matrix Monte Carlo (TMMC) method}
\label{sec:tmmc}

@@

Histogram methods typically rely on knowledge of the minimum and
maximum energies of the system in question \fixme{how?}, which are
unknown for the square-well fluid. Worse still, some square-well fluid
energies have such a low density of states that one cannot reasonably
expect to ever observe them via Monte Carlo, biased or otherwise. As
such, we need to identify an energy range of interest, bounded by
minimum and maximum ``important'' energies. \fixme{Rewrite, expand}

>>>>

\renewcommand{\title}{Optimizing Monte Carlo state space exploration
  of the square-well fluid}
\renewcommand{\author}{Michael A. Perlin}
\newcommand{\institution}{Oregon State University}
\newcommand{\department}{Department of Physics}
\newcommand{\supervisor}{Dr. David Roundy}
\renewcommand{\date}{21 April 2015}

@@

We wish to identify and develop efficient numerical methods for
determining thermodynamic properties of the square-well fluid in order
to test square-well density functional theories. The liquid-vapor
phase interface is an interesting regime for testing density
functional theories, but unbiased Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low (i.e. liquid-state) temperatures. Several
generic Monte Carlo ``histogram methods'' for collecting statistics on
low energy system states have been developed \fixme{by whom?}, but
little or no literature exists on their systematic comparison and
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model). We propose to implement these
methods in Monte Carlo simulations of the square-well fluid, determine
appropriate performance metrics for their comparison, and identify
those methods which are most efficient and effective at sampling
otherwise improbable system states.

@@

\fixme{Intro is confusing. Move these paragraphs in between sections
  on SW fluid and biased sampling?}

Understanding the behavior of fluids near the critical point, or the
point at which the distinction between liquid and gas ceases to become
well-defined, is a major challenge in the field of classical density
functional theory. Studying the critical point of a fluid via
simulations requires exploring the state space \fixme{remove words
  ``state space''?}  boundary between its liquid and gaseous
phases. For a generic square-well fluid to be observed in a liquid
state, the fluid must have a relatively localized distribution of the
spheres which make up the fluid. Such a distribution implies a low
energy fluid state; conversely, a low energy liquid state of a
square-well fluid otherwise observable in both liquid and gaseous
phases implies a localized distribution of spheres. The search for
liquid states is thus equivalent to the search for low energy
states. In unbiased Monte Carlo simulations, however, highly localized
distributions of spheres are extremely unlikely to occur, making such
simulations impractical for studying the square-well fluid.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.45\textwidth]{figs/square-well.pdf}
  \caption[The square-well pair potential]{The square-well pair
    potential can be used to model short range forces to first
    order. The infinite potential at $r<\sigma$ simply enforces the
    condition that spheres cannot overlap. At $r>\lambda\sigma$, the
    potential is zero, and makes no contribution to the fluid's
    internal energy.}
  \label{fig:pair_potential}
\end{figure}

The square-well (SW) fluid is a simple model used in classical density
functional theories to capture low order effects of short-range
attractive forces, such as the van der Waals force. The fluid is
composed of spheres with diameter $\sigma$ which have a pair potential
\begin{align}
  v_{sw}\p{\vec r}=\left\{
    \begin{array}{ll}
      \infty & \abs{\vec r}<\sigma \\
      -\epsilon & \sigma<\abs{\vec r}<\lambda\sigma \\
      0 & \abs{\vec r}>\lambda\sigma
    \end{array}
  \right., \label{eq:pair_potential}
\end{align}
also shown graphically in Figure \ref{fig:pair_potential}, where the
parameters $\lambda$ and $\epsilon$ are referred to as the well width
and depth, respectively. The first ($\abs{\vec r}<\sigma$) part of
this potential forbids spheres from overlapping, whereas the second
($\sigma<\abs{\vec r}<\lambda\sigma$) associates an energy $-\epsilon$
with each pair of spheres whose centers are within distance of
$\lambda\sigma$ of each other (where typically
$\lambda\in\sp{1.05,3}$). The net potential energy of the square-well
fluid is thus
\begin{align}
  E=\sum_{i<j}v_{sw}\p{\vec r_i-\vec r_j},
  \label{eq:internal_energy}
\end{align}
where $\vec r_i$ is the position of the $i$-th sphere. As the
potential energy $E$ is the primary form of energy concerning us in
this paper, we will refer to $E$ as simply the ``energy'' of the
fluid. An important feature of the square-well fluid is that its
energy is always an integer multiple of the well depth. A homogeneous
square-well fluid is thus uniquely identified by its well width
$\lambda$ and filling fraction $\eta$ (i.e. the proportion of space
filled by spheres; a dimensionless density), as all other properties
can be normalized to the natural energy scale $\epsilon$ and length
scale $\sigma$. In practice, our simulation codes use dimensionless
energies $E/\epsilon$, temperatures $kT/\epsilon$, and distances
$r/\sigma$.

@@

While model systems are powerful tools for understanding complex
physical systems, they do not themselves exist in the real
world. Consequently, direct experimental tests of theories for model
system (e.g. square-well density functional theories) are not
possible. For this reason, model system theories are commonly tested
against Monte Carlo simulations. Proper implementation of Monte Carlo
methods to study completely characterized systems ensures that
statistical results from simulations converge on the exact properties
of the simulated system in the infinite simulation time
limit\fixme{explain. cite?}. Furthermore, uncertainties in quantities
computed via Monte Carlo simulations are typically well-defined,
monotonically decreasing functions of simulation time, allowing one to
run simulations to the desired level of accuracy.

@@

Algorithm \ref{alg:MC} provides a sketch of unbiased Monte Carlo fluid
simulations. Such an algorithm is ``unbiased'' in the sense that it
collects statistics (i.e. data) on all valid system configurations
with equal probability. Statistics whose collection time scales as
$\mathcal O\p{1}$ with system size can be collected after every move
(defined in the algorithm), while statistics whose collection time
scales as $\mathcal O\p{N}$, where $N$ is the number of simulated
elements (e.g. spheres), can be collected after each iteration. In
general, collection with $\mathcal O\p{\chi}$ time scaling in $N$
should not occur more often than once every $\chi$ moves
\fixme{explain better?}, so as to ensure that scaling up simulations
does not cause them to asymptotically spend all of their time
collecting statistics, and no time actually simulating the
fluid. Collected statistics are used to find thermodynamic properties
of the simulated fluid.

In this work, we are concerned with simulating the {\emph homogeneous}
square-well fluid. To avoid edge effects resulting from fluid behavior
near a wall, we employ periodic boundary conditions. The use of a
finite cell with periodic boundary conditions suppresses all density
fluctuations on scales larger than the dimensions of the simulated
fluid cell, thereby introducing a source of error; addressing this
error, however, is outside the scope of this work. \fixme{Add
  feedback}

@@

Due to the fact that unbiased Monte Carlo simulations sample all of
state space randomly and without preference, a histogram $H\p{X}$ of
observations of some system property $X$ (e.g. energy) \fixme{other
  examples} is directly proportional to the density of states $D\p{X}$
in $X$ \fixme{explain?}. It is sometimes the case, however, that the
density of states in some region $R$ in the range of possible $X$ is
so low that it is practically impossible to sufficiently sample $R$,
that is, accumulate a statistically significant $H\p{R}$, in any
reasonable amount of time.

@@

Histogram methods provide a means to address unbiased Monte Carlo's
inability to sufficiently sample regions of parameter space with a low
density of states by introducing a bias into the otherwise random
sampling of state space. A weighting function $w$ is introduced, whose
domain is the value of some property $X\p{s}$ which depends on the
system state $s$. An additional condition is then added to step
\ref{alg:MC_move} of Algorithm \ref{alg:MC} in order to accept an
attempted move: the weights $w\sp{X\p{s_i}}$ and $w\sp{X\p{s_f}}$ of
the initial (pre-move) state $s_i$ and final (post-move) state $s_f$
are used to determine the probability $P\p{s_i\to s_f}$ of accepting
an otherwise valid move via
\begin{align}
  P\p{s_i\to s_f}=\min\set{\f{w\sp{X\p{s_f}}}{w\sp{X\p{s_i}}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\sp{X\p{s_f}}>w\sp{X\p{s_i}}$, the move
$s_i\to s_f$ is accepted; when $w\sp{X\p{s_f}}<w\sp{X\p{s_i}}$, the
ratio of these weights determines the probability of accepting the
move. Due to the fact that only ratios of weights determine
$P\p{s_i\to s_f}$, the weights $w\sp{X\p{s}}$ are scale-invariant,
meaning that their effect on simulations is unchanged by scale factors
that are constant with respect to $X\p{s}$.

@@

Employing biased Monte Carlo simulations, sketched out in Algorithm
\ref{alg:biased_MC}, allows one to construct weight functions that
favor some region of state space over others, as transitions to states
with higher weights are always accepted, whereas transitions to states
with lower weights may be rejected, artificially preventing the
simulated system from leaving interesting regions of state
space. Crucially, the bias introduced by weights can be reversed when
computing system properties from sampling statistics, as the weight of
a particular state is directly proportional to the probability bias of
that state; that is, a state with a weight of 2 will be sampled twice
as often as it would have been with a weight of 1. If we wish to
convert a histogram $H\p{X}$ of observations in a biased Monte Carlo
simulation into a numerical unnormalized density of states $\tilde
D\p{X}$ in $X$, we therefore divide the histogram by the corresponding
weights $w\p{X}$, i.e.
\begin{align}
  \tilde D\p{X}=\f{H\p{X}}{w\p{X}}.
  \label{eq:dos_unnormed}
\end{align}
The normalized density of states $D\p{X}$ in $X$ is then
\begin{align}
  D\p{X}=\f{\tilde D\p{X}}{\sum_X\tilde D\p{X}}
  =\f{H\p{X}/w\p{X}}{\sum_XH\p{X}/w\p{X}}.
  \label{eq:dos}
\end{align}
In general, the domain and shape of the weight function will depend on
the desired yields (e.g.  density of states, heat capacity) of a
simulation. A ``histogram method'' is an algorithm for determining a
weight function appropriate for a particular simulation.

@@

\fixme{Put canonical before this?}

The sort of Monte Carlo simulations described in this paper are only
capable of collecting statistics on system microstates, and functions
of microstates. For example, in simulation one might periodically
compute both the energy $E\p{s}$ and some system property $X\p{s}$,
both of which are determined by the system microstate $s$, in order to
find the mean value of $X$ at any given energy $E$,
i.e. $\bk{X}_E$. In the real world, however, information about a
system's microstate and functions of microstates (e.g. the energy $E$)
is generally is inaccessible. One therefore cannot measure $\bk{X}_E$
directly. Instead, one typically measures the dependence of
thermodynamic properties on macroscopic state variables such as
temperature, i.e. $\bk{X}_T$.

The temperature dependence of $\bk{X}_T$ can be expressed in terms of
the density of states $D\p{E}$ and the expectation value $\bk{X}_E$ by
\begin{align}
  \bk{X}_T=\f1{Z\p{T}}\sum_E\bk{X}_ED\p{E}e^{-E/kT},
  \label{eq:temp_dep_X_norm}
\end{align}
where the partition function $Z\p{T}$ is simply a normalization
factor, given by
\begin{align}
  Z\p{T}=\sum_ED\p{E}e^{-E/kT}.
\end{align}
To reduce redundant computations and numerical error in
implementation, we will use a partition function with an unnormalized
density of states,
\begin{align}
  \tilde Z\p{T}=\sum_E\tilde D\p{E}e^{-E/kT},
\end{align}
in terms of which
\begin{align}
  \bk{X}_T=\f1{\tilde Z\p{T}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT}.
  \label{eq:temp_dep_X}
\end{align}
Sufficient accumulation of statistics on $\bk{X}_E$ at all available
energies thus in principle allows one to determine $\bk{X}_T$ for any
$T$. In practice, the density of states can fall off so quickly with
energy that some range of allowable energies is practically
inaccessible via Monte Carlo simulations, biased or otherwise. In such
a case, computing $\bk{X}_T$ to a reasonable degree of accuracy
requires sufficiently sampling the energies at which $\tilde
D\p{E}e^{-E/kT}$ dominates the sum in (\ref{eq:temp_dep_X}).

@@

@@

A standard construction of the weight array $w\p{E}$ for what is
called a canonical Monte Carlo simulation involves \fixme{better word}
choosing a particular temperature $T_0$, and using weights
proportional to the Boltzmann factor at that temperature, i.e.
\begin{align}
  w\p{E}=e^{-E/kT_0},
\end{align}
where there is no reason to normalize $w\p{E}$ due to the fact that
only ratios of weights, as per (\ref{eq:move_prob}), are ever used in
simulation. The fact that only ratios of weights are used in
simulation makes $w\p{E}$ scale-invariant.

The partition function at $T=T_0$ for simulations with canonical
weights is
\begin{align}
  \tilde Z\p{T_0}=\sum_E\tilde D\p{E}e^{-E/kT}
  =\sum_EH\p{E}/w\p{E}~e^{-E/kT} =\sum_EH\p{E},
  \label{Z_canonical}
\end{align}
and the value of a thermodynamic property $X\p{T_0}$
\begin{align}
  X\p{T_0}=\f1{\tilde Z\p{T_0}}\sum_E\bk{X}_E\tilde D\p{E}e^{-E/kT_0}
  =\f{\sum_E\bk{X}_EH\p{E}}{\sum_EH\p{E}}.
  \label{eq:X_canonical}
\end{align}
The simplifications in in (\ref{Z_canonical}) and
(\ref{eq:X_canonical}), which have no explicit dependence on $T_0$,
occur because canonical Monte Carlo simulations sample energies in
proportion to the distribution (over energy) of microstates at a
temperature of $T_0$.  Canonical Monte Carlo simulations can therefore
fail to sufficiently sample energies which are important
(i.e. energies with a non-negligible state probability density) at
different temperatures. As a consequence, such simulations should not
be used to determine properties $X\p{T\ne T_0}$, and are thus referred
to as ``fixed temperature'' simulations.

Though canonical Monte Carlo is simple to implement, its inability to
investigate a system at more than one temperature at a time is a
disadvantage for determining the temperature dependence of system
properties. In order to find the behavior of $\bk{X}_T$, one must run
many simulations at discrete temperature intervals; each such
simulation will yield one sample (e.g. one data point) of $\bk{X}_T$.

@@

For all but the most trivial or well-studied systems, however, the
density of states $D\p{E}$ is not known prior to simulation, and is in
fact one of the system properties which a simulation is intended to
determine. Nonetheless, this toy method has pedagogical value by
enabling discussion of ``optimal'' weights generate a flat
histogram. All algorithms to construct $w\p{E}$ with the aim of
producing a flat histogram should therefore converge on the same
weight array, given by (\ref{eq:flat_weights}).

@@

\begin{algorithm}[H]
  \caption{Finding weights by optimizing energy walker currents}
  \label{alg:optimized_ensemble}
\end{algorithm}
uu
\subsection{The transition matrix Monte Carlo (TMMC) method}
\label{sec:tmmc}

@@

Histogram methods typically rely on knowledge of the minimum and
maximum energies of the system in question \fixme{how?}, which are
unknown for the square-well fluid. Worse still, some square-well fluid
energies have such an incredibly low density of states that one cannot
reasonably expect to ever observe them via Monte Carlo, biased or
otherwise. As such, we need to identify an energy range of interest,
bounded by minimum and maximum ``important'' energies. \fixme{Rewrite,
  expand}
